# Immutability Principles 

```{r}
#| label: setup
#| include: false
#| output: false
#| echo: false
# set the working directory
library(here)
here::i_am("R/Exercises/02_immutability.qmd")

# quarto render options
options(htmltools.dir.version = FALSE)

library(knitr)
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  out.width = "100%",
  fig.align = "center",
  fig.width = 6,
  fig.asp = 0.618,
  fig.show = "hold"
)

source(
  here::here(
    "R/set_r_theme.R"
  )
)
```

## Introduction

**Immutability** is a fundamental principle for reproducible research. It means **"never changing the original"** - whether that's raw data files, R objects, or data columns.

Think of immutability as creating a **clear audit trail** of all your data transformations. Instead of modifying existing data, you create new versions with documented changes.

```{r}
#| label: libraries
#| include: true
#| output: false
# Load required packages
library(tidyverse)
library(palmerpenguins) # For penguin data
library(here) # For file paths
```

### Why Immutability Matters in Ecology

1. **Preserve raw data**: Original field measurements should never be altered
2. **Enable reproduction**: Others can trace your exact steps
3. **Prevent errors**: No risk of accidentally overwriting important data
4. **Version control**: Clear history of all transformations

---

## 1. File-Level Immutability: Protecting Raw Data 📁

### The Golden Rule: Never Edit Raw Data Files

```{r}
#| label: file-immutability-bad-good
#| eval: false
#| echo: true
# ❌ BAD: Opening raw data file in Excel and "fixing" values
# This destroys reproducibility and creates no record of changes

# ✅ GOOD: Keep raw data in Data/Input/Raw/ and process in code
# All changes are documented and reproducible
```

### Exercise 1.1: Reading Raw Data

Let's start with the penguin data as our "raw" dataset:

```{r}
#| label: read-raw-data
# Treat penguins as our raw field data
data_penguin_raw <-
  palmerpenguins::penguins

readr::write_csv(
  data_penguin_raw,
  file = here::here(
    "Data/Input/penguins_raw.csv"
  )
)
```

**❌ Bad Practice:**

Keep saving over the same file:

```{r}
#| label: file-bad-practice
#| eval: false
data_penguin_raw <-
  readr::read_csv(
    here::here("Data/Input/penguins_raw.csv")
  )

data_penguin_changed <-
  data_penguin_raw %>%
  dplyr::filter(!is.na(body_mass_g))

data_penguin_changed %>%
  readr::write_csv(
    here::here("Data/Input/penguins_raw.csv")
  )
```

**✅ Good Practice:**

Save as RDS in Data/Processed/:

```{r}
#| label: file-good-practice
#| eval: false
data_penguin_raw <-
  readr::read_csv(
    here::here("Data/Input/penguins_raw.csv")
  )

data_penguin_changed <-
  data_penguin_raw %>%
  dplyr::filter(!is.na(body_mass_g))

data_penguin_changed %>%
  readr::write_rds(
    here::here("Data/Processed/penguins_processed.rds")
  )
```

## 2. Object-Level Immutability: Creating New Instead of Modifying 🔄

### The Principle: Don't Overwrite R Objects

**❌ Bad Practice (Mutable):**
```{r}
#| label: bad-practice-demo
#| eval: false
#| echo: true

# This modifies the original object - BAD!
data_penguins <-
  palmerpenguins::penguins

data_penguins <-
  data_penguins[!is.na(data_penguins$body_mass_g), ]

data_penguins <-
  data_penguins[data_penguins$year >= 2008, ]

data_penguins$body_mass_kg <-
  data_penguins$body_mass_g / 1000

# Problem: We've lost the original data and can't trace changes!
```

**✅ Good Practice (Immutable):**
```{r}
#| label: good-practice-demo
# Create new objects for each transformation - GOOD!
data_penguins_no_na <-
  palmerpenguins::penguins %>%
  dplyr::filter(!is.na(body_mass_g))

data_penguins_recent <-
  data_penguins_no_na %>%
  dplyr::filter(year >= 2008)

data_penguins_with_kg <-
  data_penguins_recent %>%
  dplyr::mutate(body_mass_kg = body_mass_g / 1000)

# We can always go back to previous steps!
nrow(data_penguins_no_na)

nrow(data_penguins_recent)

names(data_penguins_with_kg)
```

### Exercise 2.1: Immutable Data Cleaning

Let's clean some messy ecological data the immutable way:

```{r}
#| label: messy-data-creation
# Create some "messy" field data (simulating real-world issues)
data_messy <-
  tibble::tibble(
    site_id = c("SITE001", "SITE002", "SITE003", "SITE001", "SITE002"),
    species = c("Betula nana", "Salix glauca", "NA", "Vaccinium", "Salix glauca"),
    count = c(45, 32, -999, 67, 28),
    temperature_c = c(15.5, NA, 18.1, 12.3, 16.8),
    notes = c("healthy", "some damage", "missing", "flowering", ""),
    date = c("2023-07-01", "2023-07-02", "2023-07-03", "2023-07-04", "2023-07-05")
  )
data_messy
```

Now let's clean it step by step, creating new objects:

```{r}
#| label: immutable-cleaning
# Step 1: Remove invalid counts (keeping original object intact)
data_valid_counts <-
  data_messy %>%
  dplyr::filter(count >= 0)

nrow(data_valid_counts)

# Step 2: Fix species names (creating new object)
data_clean_species <-
  data_valid_counts %>%
  dplyr::mutate(
    species_clean = dplyr::case_when(
      species == "NA" ~ NA_character_,
      species == "Vaccinium" ~ "Vaccinium uliginosum",
      TRUE ~ species
    )
  )

data_clean_species %>%
  dplyr::select(species, species_clean)

# Step 3: Remove rows with missing species (new object again)
data_complete_species <-
  data_clean_species %>%
  dplyr::filter(!is.na(species_clean))

nrow(data_complete_species)

# Step 4: Final cleaned dataset
data_cleaned <-
  data_complete_species %>%
  dplyr::select(-species) %>%
  dplyr::rename(species = species_clean) %>%
  dplyr::mutate(date = as.Date(date))

data_cleaned
```

**🧠 Notice:**

- Original `data_messy` is unchanged
- Each step creates a new object
- We can inspect any intermediate step
- Full audit trail of all transformations

---

## 3. Column-Level Immutability: Adding Not Replacing 📊

### The Principle: Create New Columns Instead of Modifying Existing Ones

**❌ Bad Practice:**
```{r}
#| label: bad-column-practice
#| eval: false
#| echo: true

data_penguin <-
  palmerpenguins::penguins

# This overwrites the original column - BAD!
data_penguin$bill_length_mm <- data_penguin$bill_length_mm * 10 # Convert to cm?

data_penguin$body_mass_g <- log(data_penguin$body_mass_g) # Log transform?

# Problem: Original measurements are lost forever!
```

**✅ Good Practice:**
```{r}
#| label: good-column-practice
# Create new columns with descriptive names - GOOD!
data_penguin_measurements <-
  palmerpenguins::penguins %>%
  dplyr::filter(
    !is.na(bill_length_mm),
    !is.na(body_mass_g)
  ) %>%
  dplyr::mutate(
    # Keep original, add new
    bill_length_cm = bill_length_mm / 10,
    body_mass_log = log(body_mass_g),
    # Create derived measurements
    bill_ratio = bill_length_mm / bill_depth_mm,
    body_condition = body_mass_g / flipper_length_mm
  )

# Original data preserved, new insights available
data_penguin_measurements %>%
  dplyr::select(
    bill_length_mm, bill_length_cm,
    body_mass_g, body_mass_log,
    bill_ratio, body_condition
  ) %>%
  head()
```

---

## 4. Common Immutability Violations and Solutions ⚠️

### Exercise 5.1: Identifying Problems

**❌ Problem 1: Overwriting during exploration**
```{r}
#| label: problem-1
#| eval: false
#| echo: true

# BAD: Repeatedly modifying the same object
data <- read.csv("my_data.csv")
data <- data[data$year > 2020, ] # Lost pre-2020 data
data <- data[complete.cases(data), ] # Lost NA pattern info
data$value <- log(data$value) # Lost original scale
```

**✅ Solution 1: Create exploration pipeline**
```{r}
#| label: solution-1
# GOOD: Preserve each exploration step
data_raw <-
  palmerpenguins::penguins

# Step-by-step exploration
data_recent <-
  data_raw %>%
  dplyr::filter(year > 2007)

data_complete <-
  data_recent %>%
  dplyr::filter(complete.cases(.))

data_log_transformed <-
  data_complete %>%
  dplyr::mutate(
    body_mass_log = log(body_mass_g),
    bill_length_log = log(bill_length_mm)
  )

print(paste("Raw data:", nrow(data_raw), "rows"))
print(paste("Recent data:", nrow(data_recent), "rows"))
print(paste("Complete data:", nrow(data_complete), "rows"))
print(paste("Final data:", nrow(data_log_transformed), "rows"))
```

**🧠 Key Immutability Features:**

- Raw data never modified
- Each analysis step creates new objects  
- Original measurements preserved alongside derived ones
- Can return to any step for verification
- Clear audit trail of all transformations

---

## 5. Practice Exercises 💪

### Exercise A: Fix This Mutable Analysis

The following code violates immutability principles. Rewrite it properly:

```{r}
#| label: practice-a-bad
#| eval: false
#| echo: true

# ❌ BAD CODE - Fix this!
data <- palmerpenguins::penguins
data <- data[!is.na(data$body_mass_g), ]
data$body_mass_g <- scale(data$body_mass_g)
data <- data[data$species == "Adelie", ]
data$bill_length_mm <- log(data$bill_length_mm)
summary(data)
```

**Your task:** Rewrite using immutability principles:

```{r}
#| label: practice-a-solution
# ✅ Your solution here:
```

### Exercise B: Design an Immutable Workflow

Create an immutable analysis workflow for this research question:

**Question:** How does the relationship between flipper length and body mass change with climate (represented by year) in each penguin species?

```{r}
#| label: practice-b
# Your immutable workflow here:

# Step 1: Define raw data

# Step 2: Quality control

# Step 3: Add climate proxy and morphometric indices

# Step 4: Species-specific temporal analysis
```
